Gap formation and stability in non-isothermal protoplanetary discs

### 2014-12-14 paper/results.tex
final edit by MKL

\section{Growth of non-axisymmetric modes without the influence of the
  planet}\label{linear1} 
In this section, the planet is introduced at $t=20P_0$ and 
its potential switched on over $10P_0$. At $t=30P_0$ we switch off the
planet potential and azimuthally average the surface density, energy
and velocity fields. (At this point the planet has carved a partial
gap and the RWI has not yet occured.)
 We then perturb the surface density for $r>r_p$ and continue to 
evolve the disc. We impose  sinusoidal perturbations with 
azimuthal wavenumbers $m\in[1,10]$ and random amplitudes within $\pm 0.01$
 times the local surface density. 
This procedure allows us to analyse the growth of 
non-axisymmetric modes associated with the gap, but without
complications from non-axisymmetry arising directly from disc-planet
interaction (i.e. planet-induced wakes).

Note that these `planet-off' simulations are not linear stability
calculations since the cooling term in our energy equation
restores the initial temperature profile corresponding to $H/r=0.05$,
rather than the heated gap edge. However, we will 
examine a nearly adiabatic simulation in \S\ref{adiabatic_section},
which is closer to a proper linear problem.

Simulations here employ a resolution of $(N_r,N_{\phi})=(1024,2048)$
with open boundaries at $r=r_\mathrm{in}$ and
$r_\mathrm{out}=25r_\mathrm{in}$. We compare
cases with $\tilde{\beta}=0.1,1,10$ corresponding to fast, moderate,
and slowly cooled discs.

@@

Increasing $\tilde\beta$ leads to higher disc aspect ratios $h=H/r$,
i.e. higher temperatures. Heating mostly occur at the gap edges
due to planet-induced spiral shocks. Increasing the cooling timescale 
implies that this heat is retained in the disc. In the inviscid limit the gap
opening condition is $r_h\gtrsim H$ or $q\gtrsim 3h^3$
\citep{crida06}, which indicates that for hotter discs (higher
$h$), it becomes more difficult for a planet of fixed $q$ to open a
gap. This explains the shallower gaps in surface density when
$\tilde{\beta}$ is increased.

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0.5cm
    2cm 0cm 0cm]{figures/compare_sigma}
  \includegraphics[width=\linewidth,clip=true,trim=0.5cm
    2cm 0cm 1cm]{figures/compare_aspectratio}
\includegraphics[width=\linewidth,clip=true,trim=0.5cm
    0.5cm 0cm 1cm]{figures/compare_gvortensity}
  \caption{Azimuthally averaged gap profiles at $t=30P_0$ for the
    intial partial gap opened before instability emerges for fast
    (solid, $\tilde{\beta}=0.1$), moderate
    (dashed,  $\tilde{\beta}=1$), and slow cooling (dashed-dot,
    $\tilde{\beta}=10$). The relative surface density 
    perturbation (top), disc aspect ratio (middle) and generalized
    vortensity perturbation (bottom) are shown. \label{intial1D}}  
\end{figure}

@@

\subsection{Axisymmetric stability}
The intial planet-disc interaction form bumps and grooves in the gap profiles
which can potentially be unstable due to axisymmetric instabilities. The
generalised local axisymmetric stability condition is the Solberg-Hoiland
criterion,  
\begin{align}
  \kappa^2+N^2 \geq 0 
\end{align}
where
\begin{align}
 N^2=\frac{1}{\Sigma} \frac{\partial P}{\partial r}
 \left(\frac{1}{\Sigma} \frac{\partial \Sigma}{\partial
     r}-\frac{1}{\gamma P} \frac{\partial P}{\partial r}  \right) 
\end{align}
is the square of the Brunt-V\"ais\"al\"a frequency. 
At the outer gap edge $r=r_p+2.5r_h$,  where the RWI is excited
(see below), we find $\kappa^2 + N^2$ reaches local minimum with a value
$\sim 0.002$ (code units) for all $\tilde\beta$. The
Brunt-V\"ais\"al\"a frequency is $N\sim 10^{-5}$. 
{\bf Brunt frequency the same for all cooling?}
The Solberg-Hoiland criteria is similary satisfied for the entire disc throughout the 
simulations. {\bf i assume this means the azimuthally averaged
  values. do you ever get $N^2<0$ anywhere in the 2D disc?}  
Thus for all values of $\tilde\beta$ the planet-induced gaps are
stable to axisymmertic instabilities.

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=1.2cm
  0cm 0cm 0cm]{figures/linear_stability}
  \caption{Evolution of azimuthal Fourier modes of disc surface
    density, non-dimenionlised by the initial axisymmetric component 
    $\Sigma_0(t=0)$, for the
    `planet-off' simulation with $\tilde{\beta}=10$. Colours correspond
    to different $m$ values. The $m=3$ component is the fastest growing
    mode during linear growth and has a corresponding
    $\gamma=0.017\Omega(r_p)$.\label{linearmodes}}
  % {\bf is the $\Sigma_0$ at t=0?} }  
\end{figure}

@@

These `planet-off' simulations show that gap edges become more stable with
longer cooling times. This is expected because larger $\tilde{\beta}$
results in hotter gap profiles at $t=30P_0$ with less pronounced
generalized vortensity minima. Stabilisation with increased
cooling time is therefore due to a smoother basic state for the
instability, as it is more difficult for the planet to open a gap if
the disc is allowed to heat up.

\begin{figure}
  \centering
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/analysis_gvortensity50}
  }
\hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/analysis_gvortensity67}
  }
\hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/analysis_gvortensity75}
  }  
  \caption{Generalized vortensity perturbation (relative to $t=0$) for
    cases of $\tilde{\beta}=0.1,1,10$ (left,middle,right) during
    the growth of non-axisymmetric modes. The planet potential has
    been switched off.  The number of vortices
    decrease as $\tilde{\beta}$ increases. Note that snapshots are
    taken later for increasing $\tilde{\beta}$ because it takes longer
    for the vortices to grow and become visible with increasing cooling time. 
    \label{2Dlinear} 
  } 
\end{figure}

@@

The above `planet-off' simulations are not formally linear
stability calculations, because the cooling time is comparable or shorter
than the instability growth time, $t_c\lesssim\gamma^{-1}$.  
Thus the disk cools back to its intial temperature corresponding to
$h=0.05$ before or during the instability growth, so we do not 
have a steady basic state to formulate a standard linear stability 
problem.

In order to capture the effect of a heated gap edge, we ran a simulation with 
$\tilde{\beta}=100$, corresponding to an almost adiabtaic disc.  
In this simulation the cooling rate is slow enough that the gap 
temperature profile (e.g. middle panel of Fig. \ref{intial1D}) changes
only marginally over the instability growth timescale.

We find very similar gap profiles and mode growth rates for
$\tilde{\beta}=100$ as with $\tilde{\beta}=10$. The disc heats up to
values $h\simeq0.06$ in the nearly adiabatic case. This is close to
the original temperature of $h=0.05$, so linear growth rates are not expected
to change significantly \citep{li00}.

According to \cite{li00}, increasing $h$ increases linear growth rates
of the RWI because it is pressure-driven. However, in the case 
of disc-planet interaction, increasing $h$ has a stabilizing effect
through the gap profile because it results in smoother gap
edges. The fact that we observe smaller growth rates as $h$ is
increased indicates that for planetary gaps, the importance of $h$ on
the \emph{linear} RWI is through setting up the gap profile, i.e. basic
state for the instability (as opposed to the linear response).

\subsection{Long term evolution} \label{nonlinearplanetoff} 
{\bf typical rossby numbers for these `planet-off' vortices? do
  stronger vortices (beta=0.1) dissipate faster?
}

We also extended these `planet-off' simulations into the non-linear
regime. After the linear growth phase of the vortices, vortex merging
takes hold on timescales of up to $150P_0$, until there is one vortex
left. We find the vortex merging
time is depedendent on the growth rates of the modes and saturation
timescales, with the slowest growing modes in $\tilde\beta=10$ taking
the longest to merge.

Fig.\ref{planetofflifetimeplot} shows evolution of the $m=1$ surface
density component, which represents the amplitude of the post-merger
single vortex. For completeness we also ran intermediate cases with
$\tilde{\beta}=0.5$ and $5.0$. The vortices simply decay on a
timescale of $O(10^3)$ orbits. We will see in the next section that
this behaviour is very different to when the planet potential is kept
on. Interestingly, though, vortices arising from the stronger
instabilities (shorter cooling times or lower $\tilde{\beta}$) decays
faster.

@@

\subsection{Generic evolution} 
The linear growth of the RWI and vortex-formation is followed by 
vortex merging. We now find merging timescales independent of
$\tilde\beta$, and by $60P_o$ only one vortex remains. 
The evolution of the amplitude of the $m=1$ surface density component,
averaged over $r-r_p\in[2,10]r_h$, is shown in Fig.~\ref{lifetimeplot} 
for different $\tilde\beta$.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0.5cm
  0cm 0cm 1cm]{figures/longterm_stability}
  \caption{Evolution of the $m=1$ surface density component,
    non-dimenionlised by the initial $m=0$ component, for long term
    simulations with the planet potential kept on.\label{lifetimeplot}}   
\end{figure}

In all cases the system remains in a quasi-steady state for
$\gtrsim800P_0$ with a single vortex circulating 
the outer gap edge at the local Keplerian  
frequency. Fig. \ref{Vortex2D} shows a typical plot of the relative 
surface density perturbation in this state. During this stage, the 
vortex intensifies. As the vortex forms it has an initial 
Rossby number $Ro\approx-0.1$ for all $\tilde\beta$
(indicating anti-cyclonic 
motion). As the $m=1$ amplitude grows the Rossby number also increases
in magnitude, 
 reaching a characteristic minimum value of $Ro\approx-0.35$ to $-0.45$ for
all cooling rates. 
{\bf Maybe useful to have a Rossby number v.s. time plot to complement
  mode amplitude plot. 
}

We find the vortices can become significantly over-dense. 
The surface density perturbation 
can reach values of $\Delta\Sigma/\Sigma_0 \sim 7$ at the center of vortices
for all cases of $\tilde\beta$ in quasi-steady state, and
was found as high as $\Delta\Sigma/\Sigma_0 \sim 9$ for the 
$\tilde\beta=10$ case. %despite beta=10 having the weakest linear
% instability 
{\bf any trend between maximum over-density achieved and cooling?} 
This large increase in surface density is due to  
vortex growth, since there is continuous generation of vorticity by
planet-disc interaction. This is supported by the observation that in
the previous simulations without the planet, the amplitude of
the post-merger vortex does not grow (Fig
\ref{planetofflifetimeplot}).  We remark that the increase in the
vortex surface density directly due to the pile-up of material at the
gap edge (because of gap-opening) is small: the average surface density
perturbation at the outer gap edge, $r\sim r_{p}+2.5r_h$, is 
$\langle\Delta\Sigma/\Sigma_0\rangle_\phi<1$.  
{\bf may be more appropriate to quote the azimuthally averaged
  perturbation at the location of the vortex (5-6 Hill radii from the
  planet?)}

Fig. \ref{lifetimeplot} shows that the duration of the quasi-steady state
varies with the cooling rate: for 
$\tilde{\beta}=0.1$ and $10$, the vortex amplitude begins to decay around
$t\sim900P_0$, while for $\tilde{\beta}=1,\,5$ the decay begins at 
$t\sim1200P_0$. This non-monotonic dependence suggests that there
exists an optimal cooling rate to maximise the vortex lifetime. We
will discuss this issue futher in
\S\ref{lifetime_discuss}. Notice the decay timescale decreases with
increasing cooling time: for $\tilde{\beta}=0.1$ it takes $\sim400P_0$
whereas for $\tilde{\beta}=10$ it takes $\sim 100P_0$ for the $m=1$
amplitude to decay significantly after reaching maximum.

{\bf question: vortex amplitude plot suggest just after linear growth,
  vortex amplitudes are smaller with increasing beta, but later on in
  the quasi-steady state, this trend reverses?  
}
{\bf Perhaps a 2D snapshot of what the gap edge looks like after
  vortex death (like the one in the poster). defer definition of
  vortex lifetime to later, when we discuss it. 
}

@@

@@

In quasi-steady ($t=700P_0$) the vortex is elongated, but becomes
more compact as it grows before decaying. 
{\bf rough estimate of
  vortex aspect-ratio in the three snapshots? 
  (length/width, by inspection is ok)}  
Notice in Fig. \ref{shockplot} the appearence of wakes extending from
either side of the vortex at $t=1300P_0$. These 
wakes correlate with large gradients in surface density (bottom
panel), and are first seen in the later half of the quasi-steady
state. We find the time at which 
the vortex begins to decay coincides with the emergence of these
wakes. {\bf true statement?}

% As the vortices grow in intensity wakes are seen to extend from them. 
During quasi-steady state the average value of the surface density gradient
along the wakes is $|\nabla\Sigma/\Sigma| \sim 4 $.{\bf this quantity
  has dimensions of 1/length, preferable to quote something
  dimensionless, or say `in code units'}  
Just before the $m=1$ amplitude begins to decrease, we observe this quantity
sharply increases to $ \sim 6 $, and remains around this value
until the vortex dies out, at which point the associated Rossby number
quickly approaches zero.

\begin{figure}
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock1}
  }
  \hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock2}
  }
  \hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock3}
  } \\[-0.98cm]
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock4}
  }
  \hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock5}
  }
  \hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock6}
  } 
  \caption{The vortex in the case with $\tilde{\beta}=1$
    during quasi-steady state (left), at maximum amplitude of the
    $m=1$ surface density component, 
    (middle), and just after the decay in the $m=1$ amplitude
    (right). The surface density perturbation
    (top) and the associated surface density gradient (bottom) are
    shown. This plot can considered in conjunction with
    Fig. \ref{lifetimeplot}. 
    \label{shockplot}}
  % Large surface density gradients become prominant
  % around vortex during dissipation} 
\end{figure}

We also measured large increases in the Mach number near
the vortex as it reaches maximum $m=1$ amplitude and begins to decay. 
Figure \ref{machplot} plots the Mach number $M=|\bm{v} -
\bm{v}_\mathrm{vor}|/c_s$, where 
$\bm{v}_\mathrm{vor}$ corresponds to the bulk velocity of the vortex
around the disc. Values in Fig. \ref{machplot} have been averaged over
a region within $2H$ of the vortex centre. 
During the quasi-steady state the Mach number increase
steadily, and for all cases $M$ reaches a maximum at 
% with larger growth rate for higher $\tilde\beta$.
a time coincident with the start of vortex decay. {\bf check if
  statement is true}

Putting the above observations together, we suggest the decay
in the vortex amplitude is due to shock formation by the
vortex. When the vortex reaches large amplitude, it begins
to induce shocks in the surrounding fluid, as supported by the
increase in Mach number and the appearence of wakes with large surface
density gradients. The vortex may lose energy through shock
dissipation. In addition, a strong vortex (or shock formation) 
can smooth out the gap structure that originally gave rise to the
RWI, which would oppose vortex growth. We examine this below.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/mach}
  \caption{Running-time average {\bf (?)} of the Mach number relative to
    the vortex, averaged over a region within $2H$ of the  
    vortex centre. This plot can be compared to the evolution of the
    vortex amplitude shown in Fig. \ref{lifetimeplot}.
    \label{machplot}}
\end{figure}

>>>>

\section{Growth of non-axisymmetric modes without the influence of the
  planet}\label{linear1} 
In this section, the planet is introduced at $t=20P_0$ and 
its potential is switched on over $10P_0$. At $t=30P_0$ we switch off the
planet potential and azimuthally average the surface density, energy
and velocity fields. (At this point the planet has carved a partial
gap and the RWI has not yet occurred.) Effectively, we initialise the
disc with a gap profile. 
We then perturb the surface density in the outer disc ($r>r_p$) and continue to 
evolve the disc. We impose  sinusoidal perturbations with 
azimuthal wavenumbers $m\in[1,10]$ and random amplitudes within $\pm 0.01$
times the local surface density. 
This procedure allows us to analyse the growth of 
non-axisymmetric modes associated with the gap, but without
complications from non-axisymmetry arising directly from disc-planet
interaction (i.e. planet-induced wakes).

Note that these `planet-off' simulations are not linear stability
calculations because the cooling term in our energy equation
restores the initial temperature profile corresponding to constant
$H/r=0.05$, rather than the heated gap edge. However, we will 
examine a nearly adiabatic simulation in \S\ref{adiabatic_section},
which is closer to a proper linear problem.

Simulations here employ a resolution of $(N_r,N_{\phi})=(1024,2048)$
with open boundaries at $r=r_\mathrm{in}$ and
$r_\mathrm{out}=25r_\mathrm{in}$. We compare cases with
$\tilde{\beta}=0.1,1,10$ corresponding to fast, moderately, 
and slowly cooled discs.

@@

Increasing $\tilde\beta$ leads to higher disc aspect ratios $h=H/r$,
i.e. higher temperatures. Heating mostly occurs at the gap edges
due to planet-induced spiral shocks. Increasing the cooling timescale 
implies that this heat is retained in the disc. In the inviscid limit the gap
opening condition is $r_h\gtrsim H$ or $q\gtrsim 3h^3$
\citep{crida06}, which indicates that for hotter discs (higher
$h$), it becomes more difficult for a planet of fixed $q$ to open a
gap. This explains the shallower gaps in surface density when
$\tilde{\beta}$ is increased.

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0.5cm
    2cm 0cm 0cm]{figures/compare_sigma}
  \includegraphics[width=\linewidth,clip=true,trim=0.5cm
    2cm 0cm 1cm]{figures/compare_aspectratio}
\includegraphics[width=\linewidth,clip=true,trim=0.5cm
    0.5cm 0cm 1cm]{figures/compare_gvortensity}
  \caption{Azimuthally averaged gap profiles at $t=30P_0$ for the
    initial partial gap opened before instability emerges for fast
    (solid, $\tilde{\beta}=0.1$), moderate
    (dashed,  $\tilde{\beta}=1$), and slow cooling (dashed-dot,
    $\tilde{\beta}=10$). The relative surface density 
    perturbation (top), disc aspect ratio (middle) and generalised
    vortensity perturbation (bottom) are shown. \label{intial1D}}  
\end{figure}

@@

\subsection{Axisymmetric stability}
The initial planet-disc interaction form bumps and grooves in the gap profiles
which can potentially be unstable due to axisymmetric instabilities. The
generalised local axisymmetric stability condition is the Solberg-Hoiland
criterion,  
\begin{align}
  \kappa^2+N^2 \geq 0 
\end{align}
where
\begin{align}
 N^2=\frac{1}{\Sigma} \frac{\partial P}{\partial r}
 \left(\frac{1}{\Sigma} \frac{\partial \Sigma}{\partial
     r}-\frac{1}{\gamma P} \frac{\partial P}{\partial r}  \right) 
\end{align}
is the square of the Brunt-V\"ais\"al\"a frequency. 
At the outer gap edge $r=r_p+2.5r_h$,  where the RWI is excited
(see below), we find $\kappa^2 + N^2$ reaches local minimum with a value
$\sim 2\times10^{-4}$ (code units) for all $\tilde\beta$. The
Brunt-V\"ais\"al\"a frequency is $N\sim 10^{-5}$, decreasing marginally with longer cooling rate. 
The Solberg-Hoiland criteria is similarly satisfied for the entire 2D disc throughout the 
simulations.
Thus for all values of $\tilde\beta$ the planet-induced gaps are
stable to axisymmertic instabilities.

% {\bf Brunt frequency the same for all cooling?:small decrease with beta,
%  not more than order of magnitude}

%  {\bf i assume this means the azimuthally averaged
%   values. do you ever get $N^2<0$ anywhere in the 2D disc?:
%  This was not for azimuthal average, below 0 inside gap but kappa+N always above 0 everywhere}

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=1.2cm
  0cm 0cm 0cm]{figures/linear_stability}
  \caption{Evolution of azimuthal Fourier modes of disc surface
    density, non-dimenionlised by the initial axisymmetric component 
    $\Sigma_0(t=0)$, for the
    `planet-off' simulation with $\tilde{\beta}=10$. Colours correspond
    to different $m$ values. The $m=3$ component is the fastest growing
    mode during linear growth with a growth rate of 
    $\gamma=0.017\Omega(r_p)$.\label{linearmodes}}
  % {\bf is the $\Sigma_0$ at t=0?} }  
\end{figure}

@@

These `planet-off' simulations show that gap edges become more stable with
longer cooling times. This is expected because larger $\tilde{\beta}$
results in hotter gap profiles at $t=30P_0$ with less pronounced
generalised vortensity minima. Stabilisation with increased
cooling time is therefore due to a smoother basic state for the
instability, as it is more difficult for the planet to open a gap if
the disc is allowed to heat up.

\begin{figure}
  \centering
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/analysis_gvortensity50}
  }
\hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/analysis_gvortensity67}
  }
\hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/analysis_gvortensity75}
  }  
  \caption{Generalised vortensity perturbation (relative to $t=0$) for
    cases of $\tilde{\beta}=0.1,1,10$ (left,middle,right) during
    the growth of non-axisymmetric modes. The planet potential has
    been switched off.  The number of vortices
    decrease as $\tilde{\beta}$ increases. Note that snapshots are
    taken later for increasing $\tilde{\beta}$ because it takes longer
    for the vortices to grow and become visible with increasing cooling time. 
    \label{2Dlinear} 
  } 
\end{figure}

@@

The above `planet-off' simulations are not formally linear
stability calculations, because the cooling time is comparable or shorter
than the instability growth time, $t_c\lesssim\gamma^{-1}$.  
Thus the disk cools back to its initial temperature corresponding to
$h=0.05$ before or during the instability growth, so we do not 
have a steady basic state to formulate a standard linear stability 
problem.

In order to capture the effect of a heated gap edge, we ran a simulation with 
$\tilde{\beta}=100$, corresponding to an almost adiabatic disc.  
In this simulation the cooling rate is slow enough that the gap 
temperature profile (e.g. middle panel of Fig. \ref{intial1D}) changes
only marginally over the instability growth timescale.

We find very similar gap profiles and mode growth rates for
$\tilde{\beta}=100$ as with $\tilde{\beta}=10$. At $t=30P_0$, the disc only heats up to
values $h\simeq0.06$ in the nearly adiabatic case. This is close to
the original temperature of $h=0.05$, so linear growth rates are not expected
to change significantly \citep{li00}.

According to \cite{li00}, increasing $h$ increases linear growth rates
of the RWI because it is pressure-driven. However, in the case 
of disc-planet interaction, increasing $h$ has a stabilizing effect
through the setting up the gap profile because it results in smoother gap
edges. The fact that we observe smaller growth rates as $h$ is
increased indicates that for planetary gaps, the importance of $h$ on
the \emph{linear} RWI is through setting up the gap profile, i.e. basic
state for the instability (as opposed to the linear response).

\subsection{Long term evolution} \label{nonlinearplanetoff} 
% {\bf typical rossby numbers for these `planet-off' vortices? do
%   stronger vortices (beta=0.1) dissipate faster?:very complicated
% maybe not worth discussing since just comparison to planet-off}

We also extended these `planet-off' simulations into the non-linear
regime. After the linear growth phase of the vortices, vortex merging
takes hold on timescales of up to $150P_0$, until there is one vortex
left. We find the vortex merging
time is dependent on the growth rates of the modes and saturation
timescales, with the slowest growing modes in $\tilde\beta=10$ taking
the longest to merge.

Fig.\ref{planetofflifetimeplot} shows evolution of the $m=1$ surface
density component, which represents the amplitude of the post-merger
single vortex. For completeness we also ran intermediate cases with
$\tilde{\beta}=0.5$ and $5.0$. The amplitude of the initially formed vortex was 
found to decrease with increased cooling rate. The vortices simply decay on a
timescale of $O(10^3)$ orbits with faster decay for stronger vortices
(which are obtained with faster cooling rates). We will see in the
next section that this behaviour is very different to when the planet
potential is kept on.

%Interestingly, though, vortices arising from the stronger
%instabilities (shorter cooling times or lower $\tilde{\beta}$) decays
%faster.

@@

\subsection{Generic evolution} 
The linear growth of the RWI and vortex-formation is followed by 
vortex merging. We now find merging timescales independent of
$\tilde\beta$, and by $60P_o$ only one vortex remains. 
The evolution of the amplitude of the $m=1$ surface density component,
averaged over $r-r_p\in[2,10]r_h$, is shown in Fig.~\ref{lifetimeplot} 
for different $\tilde\beta$. The initial, post-merger vortex amplitude
is found to be weaker for longer cooling rates (which have smaller 
linear growth rates).

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0.5cm
  0cm 0cm 1cm]{figures/longterm_stability}
  \caption{Evolution of the $m=1$ surface density component,
    non-dimenionlised by the initial $m=0$ component, for long term
    simulations with the planet potential kept on. Notice there is
    vortex growth after the initial linear growth, which
    contrasts to Fig. \ref{planetofflifetimeplot}, where vortices
    decay in the absence of continuous disc-planet interaction. 
    \label{lifetimeplot}}   
\end{figure}

In all cases the system remains in a quasi-steady state for
$\gtrsim800P_0$ with a single vortex circulating 
the outer gap edge at the local Keplerian  
frequency. Fig. \ref{Vortex2D} shows a typical plot of the relative 
surface density perturbation in this state. During this stage, the 
vortex intensifies. This is better shown in Fig. \ref{rossbyplot} as the
evolution of Rossby numbers measured at the vortex centres.  
%As the vortex forms it has an initial 
%Rossby number $Ro\approx-0.1$ for all $\tilde\beta$
%(implying anti-cyclonic 
%motion). 
The Rossby number increases in magnitude during quasi-steady state,
but the maximum $|Ro|$ is similar for all $\tilde{\beta}$: the vortex
reaches a characteristic value of $Ro\approx-0.35$ for 
$\tilde\beta=0.1$ and $Ro\approx -0.45$ for $\tilde\beta=10$.

% {\bf Maybe useful to have a Rossby number v.s. time plot to complement
% mode amplitude plot. }

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0.5cm
  0cm 0cm 1cm]{figures/rossby}
  \caption{Evolution of Rossby numbers at the centres of the vortices
    form in in discs with different cooling rates. A negative Rossby
    number implies anti-cyclonic motion. Boxcar averaging
    was used to remove contributions from the planet-induced spiral shock.\label{rossbyplot}}    
\end{figure}

We find the vortices become significantly 
over-dense. Fig. \ref{overdensity} plots the surface density perturbation measured at the
vortex centres, showing $\Delta\Sigma/\Sigma_0 \gtrsim 7$ for all
cases of $\tilde\beta$ in quasi-steady state, and 
$\mathrm{max}(\Delta\Sigma/\Sigma_0)\sim 11$ for $\tilde\beta=5$.  
The maximum over-density typically increases with longer cooling
times, despite the vortices are initially weaker at formation with
increasing $\tilde{\beta}$. The large increase in the surface density
is due to vortex growth as there is continuous generation of vorticity by
planet-disc interaction. This is supported by the observation that in
the previous simulations without the planet, the amplitude of
the post-merger vortex does not grow (Fig. 
\ref{planetofflifetimeplot}).

% We remark that the increase in the
% vortex surface density directly due to the pile-up of material at the
% gap edge (because of gap-opening) is small: the average surface density
% perturbation at vortex radial region, $r\sim r_{p}+5r_h$, is 
% $\langle\Delta\Sigma/\Sigma_0\rangle_\phi<1$.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0.5cm
  0cm 0cm 1cm]{figures/vortex_density}
  \caption{Average value of the relative surface density perturbation
    at vortex centres for various cooling
    times. \label{overdensity}}     
\end{figure}

Fig. \ref{lifetimeplot} shows that the duration of the quasi-steady state
varies with the cooling rate: for 
$\tilde{\beta}=0.1$ and $10$, the vortex amplitude begins to decay around
$t\sim900P_0$, while for $\tilde{\beta}=1,\,5$ the decay begins at 
$t\sim1200P_0$. This non-monotonic dependence suggests that there
exists an optimal cooling rate to maximise the vortex lifetime. We
will discuss this issue further in 
\S\ref{lifetime_discuss}. Notice the decay timescale decreases with
increasing cooling time: for $\tilde{\beta}=0.1$ it takes $\sim400P_0$
whereas for $\tilde{\beta}=10$ it takes $\sim 100P_0$ for the $m=1$
amplitude to decay significantly after reaching maximum.

% {\bf may be more appropriate to quote the azimuthally averaged
%   perturbation at the location of the vortex (5-6 Hill radii from the
%   planet?): yes, this was my original idea, mistyped as 2.5
% }

% {\bf any trend between maximum over-density achieved and cooling?: generally increasing with beta, max at beta=5}

% {\bf question: vortex amplitude plot suggest just after linear growth,
%   vortex amplitudes are smaller with increasing beta, but later on in
%   the quasi-steady state, this trend reverses? yes. 
% }

% {\bf Perhaps a 2D snapshot of what the gap edge looks like after
%   vortex death (like the one in the poster).: Not sure if this is figure 
% you were thinking (Fig 8)
% % defer definition of  vortex lifetime to later, when we discuss it. 
% }

% \begin{figure}
%   \includegraphics[width=.3\linewidth,height=.5\linewidth]{figures/analysis_sigma100}
%   \includegraphics[width=.3\linewidth,height=.5\linewidth]{figures/analysis_sigma130}
%   \includegraphics[width=.3\linewidth,height=.5\linewidth]{figures/analysis_sigma200}

%   \caption{2D gap profile for $\tilde\beta=1.0$ during quasisteady state (left), during dissipation (middle), and after dissipation (right).
%     \label{2dgaps}} 
% \end{figure}

@@

In quasi-steady ($t=700P_0$) the vortex is elongated with a vortex aspect ratio $ \approx 4$
, but becomes more compact approaching a ratio of 2 during its decay ($t=1310$).   
Notice in Fig. \ref{shockplot} the appearance of wakes extending from
either side of the vortex at $t=1300P_0$. These 
wakes correlate with large gradients in surface density (bottom
panel), and are first seen in the later half of the quasi-steady
state. We find the time at which 
the vortex begins to decay coincides with the emergence of these
wakes. %{\bf true statement? yes, distortion of vortex shape form wakes happens at same time}

% As the vortices grow in intensity wakes are seen to extend from them. 
During quasi-steady state the average value of the surface density gradient
along the wakes is $|w_s\nabla\Sigma/\Sigma| \sim 0.4 $, where 
$w_s\simeq0.1$ (code units) is a typical length scale of 
the surface density variation across the wake.   
Just before the $m=1$ amplitude begins to decrease, we observe this quantity
sharply increases to $ \sim 0.6 $, and remains around this value
until the vortex dies out, at which point the associated Rossby number
begins decreasing to zero
%{\bf :check}.

% {\bf rough estimate of
%   vortex aspect-ratio in the three snapshots? 
%   (length/width, by inspection is ok):trend of decreasing aspect ratio is common for all beta}  
% {\bf this quantity has dimensions of 1/length, preferable to quote something
%   dimensionless, or say `in code units': used shock width since relavant
%  distance, since small may appear to weaken value/argument}

\begin{figure}
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock1}
  }
  \hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock2}
  }
  \hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock3}
  } \\[-0.98cm]
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock4}
  }
  \hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock5}
  }
  \hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock6}
  } 
  \caption{The vortex in the case with $\tilde{\beta}=1$
    during quasi-steady state (left), start of decay 
    (middle), and just after the decay in the $m=1$ amplitude
    (right). The surface density perturbation
    (top) and the associated surface density gradient (bottom) are
    shown. This plot is to be considered in conjunction with
    Fig. \ref{lifetimeplot}. 
    \label{shockplot}}
  % Large surface density gradients become prominant
  % around vortex during dissipation} 
\end{figure}

We also measured large increases in the Mach number near
the vortex as the $m=1$ surface density amplitude reaches maximum and begins to decay. 
Fig. \ref{machplot} plots the Mach number $M=|\bm{v} -
\bm{v}_\mathrm{vor}|/c_s$, where 
$\bm{v}_\mathrm{vor}$ corresponds to the bulk velocity of the vortex
around the disc. Values in Fig. \ref{machplot} have been averaged over
a region within $2H$ of the vortex centre. 
During the quasi-steady state the Mach number increases 
steadily, and for all cases $M$ maximises about
% with larger growth rate for higher $\tilde\beta$.
$\sim 100P_0$ after the start the $m=1$ surface density
amplitude starts to decay. 
% {\bf check if  statement is true}

Putting the above observations together, we suggest that vortex decay
(in the $m=1$ surface density amplitude) is due to shock formation by
the vortex. When the vortex reaches large amplitude, it begins
to induce shocks in the surrounding fluid, as supported by the
increase in Mach number and the appearance of wakes with large surface
density gradients. The vortex may lose energy through shock
dissipation. In addition, a strong vortex (or shock formation) 
can smooth out the gap structure that originally gave rise to the
RWI, which would oppose vortex growth. We examine this below.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/mach}
  \caption{Mach number relative to
    the vortex, averaged over a region within $2H$ of the  
    vortex centre with respect to time. This plot can be compared to the evolution of the
    vortex amplitude shown in Fig. \ref{lifetimeplot}.
    \label{machplot}}
\end{figure}

\subsection{Vortex decay and gap structure}  
We find vortex decay modifies the gap
structure. Fig. \ref{gap_smoothed} shows the gap profile before and
after vortex decay for the case $\tilde{\beta}=1$. The vortex resides
around the local surface density maximum at the outer gap edge ($r\sim r_p +
6r_h$). We see that after its amplitude has decayed ($t\sim1500P_0$,   
Fig. \ref{lifetimeplot}), this local surface density maximum is also 
smoothed out.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/gapchange}
  \caption{Azimuthally averaged profiles of the relative surface
    density perturbations for  $\tilde\beta=1.0$
    before (solid) and after vortex decay
    (dashed).
    \label{gap_smoothed}} 
\end{figure}
